In the Age of Web:\\ \textnormal{\LARGE Typed Functional-First Programming Revisited}

### 2015-01-06 ml-2014-paper/paper.tex
Spellcheck

Among the ML family of languages, F\# often takes a pragmatic approach and emphasizes ease of use
and the ability to integrate with its execution environments\footnote{Historically, this applied 
to the .NET runtime, but the same applies to integrating with JavaScript in the web context.} over 
other aspects of language design. If you use the F\# langage as ML, you get most of the good 
well-known properties of ML\footnote{For example, F\# does not require type annotations when used 
as ML, but requires them when used with .NET objects.}. However, F\# leaves enough \emph{holes} that
let you use it \emph{not} as ML. This is the space that we explore in this paper.

@@

\section{Case Study: Web-based data analytics}
\label{sec:case}
In this case study, we develop a web application shown in Figure~\ref{fig:wb}, which lets the
user compare university enrolment in a number of selected countries and regions around the world. 
The resulting application runs on the client-side (as JavaScript) and fetches data dynamically 
from the World Bank \cite{data-wb-schter}.

@@

To access the university enrollment information, we first obtain a list of countries using the 
World Bank type provider from the F\# Data library \cite{fsharp-data}. The type provider exposes 
the individual countries as members of an object (the notation \lident{Country Name} is used for 
identifiers with spaces):
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{WorldBank}~=~\ident{WorldBankData}\langl\ident{Asynchronous}=\kvd{true}\rangl
 \\[0.5em]
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countries}~= \\
 \quad~\lbrack~~ \ident{data.Countries.}\lident{European Union} \\
 \qquad   \ident{data.Countries.}\lident{Czech Republic} \\
 \qquad   \ident{data.Countries.}\lident{United Kingdom} \\
 \qquad   \ident{data.Countries.}\lident{United States} ~\rbrack 
\end{array}
\end{equation*}
%
The type provider connects to the World Bank and obtains a lisf of countries at 
\emph{compile-time} and at \emph{edit-time} (when using auto-completion in an editor).
This means that the list is always up-to-date and we get a compile time error when accessing 
a country that no longer exists (a property discussed in Section~\ref{sec:tp-data}).

@@

Importing types for JavaScript libraries into the F\# type system has interesting implications, 
because the TypeScript language does not have the traditional type safety property \cite{ms-safets}.
We return to this topic in Section~\ref{sec:tp-lang}. Next, we generate checkboxes that appear on the 
right in Figure~\ref{fig:wb}:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{jQuery}~\ident{command}~=~\ident{j.jQuery.Invoke}(\ident{command}) 
 \\[0.5em]
 \kvd{let}~\ident{infos}~=~\ident{countries}~|\rang~\ident{List.map}~(\kvd{fun}~\ident{country}~\rightarrow \\
 \quad \kvd{let}~\ident{inp}~=~\ident{jQuery}(\str{<input>})\ident{.attr}(\str{type}, \str{checkbox}) \\
 \quad \ident{jQuery}(\str{\#panel})\ident{.append}(\ident{inp})\ident{.append}(\ident{country.Name}) \\
 \quad \ident{country.Name}, \ident{country.Indicators}, \ident{el}) \\
\end{array}
\end{equation*}
%
To manipulate the DOM (Document Object Model), we are using the jQuery library in a way that is
very similar to code that one would write in JavaScript. We define a helper function \ident{jQuery}
(hiding some of the complexities of the mapping) and use it to create the \str{<input>} element and
specify its attributes. Note that members like \ident{append} and \ident{attr} are standard 
jQuery patterns. The compiler sees them as ordinary object members. When writing code using F\# editors
based on the F\# Compiler Serivce \cite{fsharp-fcs}, they also appear in the auto-complete list.

@@

The main part of the sample program is a function \ident{render} that asynchronously fetches 
data for selected countries and generates a chart. To keep the code simple, we iterate over the 
\ident{infos} list from the previous section and load data for countries one by one:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{render}~()~=~\ident{async}~\{ \\
 \quad \kvd{let}~\ident{head}~=~\str{School enrollment, tertiary (\% gross)} \\
 \quad \kvd{let}~\ident{o}~=~\ident{h.HighchartsOptions}() \\
 \quad \ident{o.chart} \leftarrow \ident{h.HighchartsChartOptions}(\ident{renderTo}=\str{plc}) \\
 \quad \ident{o.title} \leftarrow \ident{h.HighchartsTitleOptions}(\ident{text}~=~\ident{head}) \\
 \quad \ident{o.series} \leftarrow \lbrack| ~ |\rbrack
 \\[0.5em]   
 \quad \kvd{for}~\ident{name},~\ident{ind},~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do}\\
 \qquad   \kvd{if}~\ident{unbox}\langl \ident{bool} \rangl~(\ident{check.is}(\str{:checked}))~\kvd{then} \\
 \qquad\quad     \kvd{let!}~\ident{v}~=~\ident{ind.}\lident{School enrollment, tertiary (\% gross)} \\
 \qquad\quad     \kvd{let}~\ident{data}~=~\ident{vals}~
                   |\rang~\ident{Seq.map}~(\kvd{fun}~(\ident{k}, \ident{v})~\rightarrow \lbrack|~\ident{number~k};~\ident{number~v} ~|\rbrack)~
                   |\rang~\ident{Array.ofSeq} \\
 \qquad\quad     \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name})) ~\}
\end{array}
\end{equation*}
%
Although the function looks like ordinary code, it is wrapped in the $\ident{async}~\{\ldots\}$ 
block, which is an F\# computation expression \cite{fsharp-zoo}. The F\# compiler performs 
de-sugaring similar to the CPS transformation and interprets keywods such as \kvd{let!} and 
\kvd{for} using special operations (monadic bind and other). The \ident{async} identifier 
determines that we are writing asynchronous workflow \cite{fsharp-async} that makes it 
possible to include non-blocking calls in the block.

@@

\subsection{Larning from the case study}

@@

In terms of programming language theory, type providers can be easily explained anecdotally. 
They change the starting point for a type-checking of a program as follows:
%
\begin{equation*}
\hspace{8em}
\begin{array}{rclll}
 \emptyset &\narrow{\vdash}& e : \tau 
    &\qquad& (\textit{classical, or closed-world})\\
 \pi~(\includegraphics[width=1em,trim=0 1.5em 0 0]{earth.png})  &\narrow{\vdash}& e : \tau
    &\qquad& (\textit{type proiders, or open-world})
\end{array}
\end{equation*}
%
Rather than starting the type-checking with an empty context, we start type-checking with a
context containing a \emph{projection} of some information from the world. In this paper, we do not
attempt to formalize type providers, so we avoid the question of what, formally, is the world here.
However, this analogy is useful for understanding how type providers work and how they can go wrong:

@@

In summary, the World Bank type provider fits a specific niche that we were exploring in the case 
study. That is, when we want to create an information analysis that accesses specific data from
an information source. With rising popularity of data journalism and Open Government Data
initiatives, we argue that this is an important problem domainHere, we first specify the country 
codes and then obtain the country objects.

@@

Type providers are often described as a technique that simplifies data access. The TypeScript
type provider shows that there is a broader range of uses, including integration between different
programming languages. Another example of type provider for language integration is the R
provider \cite{fsharp-rprovider} which imports packages and functions of the statistical 
environemnt R.

@@

\item Second, if the imported language has a more precise type system than F\#, the type 
  provider hsa to drop some of the information. This can be done safely for types in 
  contravariant positions. In the unsafe case, the runtime needs to perform dynamic checks 
  (Section~\ref{sec:conc-interop}).
\end{itemize}

@@

An interesting example of type provider for a language with a weaker type system is the
R provider \cite{fsharp-rprovider}. It solves the mismatch by using runtime reflection 
to discover R functions and their declared parameters. However, R provider also generates
additional function overloads that let the user specify additional parameters (not explicitly 
declared in the signature). To make such interoperability easier, we propsed additional 
escape mechanism \cite{fsharp-paramsdict} that provide usable syntax when static type information 
cannot be fully recovered.

@@

Using a type provider is similar to writing a code generator that turns TypeScript 
\textcolor{strclr}{\ttfamily d.ts} files into the above OCaml annotation. The main difference
is that type providers provide a smoother integration. Type providers do not produce any artefacts
(generated files) and can import entire repositories lazilly.

@@

When we write code using type providers that provide erased types (both World Bank and TypeScript), 
then the erasure happens before a quotation is captured. The following listing shows an example. 
When we write code with explicit (or implicit) quotation containing a povided method (1), the actual 
quotation (2) contains just calls to the underlying runtime library:
%
\begin{equation*}
\begin{array}{llr}
 \langl @~\ident{data.}\lident{Czech Republic}~@\rangl &\hspace{10em}& (1)\\
 \langl @~\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE}) ~@\rangl  && (2)
\end{array}
\end{equation*}
%
Thus, the FunScript component only needs to provide a client-side implementation of the underlying
operations \ident{GetCountries}, \ident{GetCountry} and \ident{AsyncGetIndicator}. This is done in the
same way as mappings for standard .NET libraries. Under the cover, \ident{AsyncGetIndicator} invokes
an AJAX request to the World Bank API. We provide more details about asynchronous execution in 
Section~\ref{sec:js-async}.

@@

\begin{itemize}
\item Numeric types including \ident{int} and \ident{float} are mapped to JavaScript numbers.
  However, the floating-point artihmetic in JavaScript differs from the one in F\# and so this changes 
  the semantics. Similarly, compilation of integer operators requires additional work (for 
  example, $1/2$ is not an integer division in JavaScript).

>>>>

Among the ML family of languages, F\# often takes a pragmatic approach and emphasizes ease of use
and the ability to integrate with its execution environments\footnote{Historically, this applied 
to the .NET runtime, but the same applies to integrating with JavaScript in the web context.} over 
other aspects of language design. If you use the F\# language as ML, you get most of the good 
well-known properties of ML\footnote{For example, F\# does not require type annotations when used 
as ML, but requires them when used with .NET objects.}. However, F\# leaves enough \emph{holes} that
let you use it \emph{not} as ML. This is the space that we explore in this paper.

@@

\section{Case Study: Web-based data analytics}
\label{sec:case}
In this case study, we develop a web application shown in Figure~\ref{fig:wb}, which lets the
user compare university enrollment in a number of selected countries and regions around the world. 
The resulting application runs on the client-side (as JavaScript) and fetches data dynamically 
from the World Bank \cite{data-wb-schter}.

@@

To access the university enrollment information, we first obtain a list of countries using the 
World Bank type provider from the F\# Data library \cite{fsharp-data}. The type provider exposes 
the individual countries as members of an object (the notation \lident{Country Name} is used for 
identifiers with spaces):
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{WorldBank}~=~\ident{WorldBankData}\langl\ident{Asynchronous}=\kvd{true}\rangl
 \\[0.5em]
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countries}~= \\
 \quad~\lbrack~~ \ident{data.Countries.}\lident{European Union} \\
 \qquad   \ident{data.Countries.}\lident{Czech Republic} \\
 \qquad   \ident{data.Countries.}\lident{United Kingdom} \\
 \qquad   \ident{data.Countries.}\lident{United States} ~\rbrack 
\end{array}
\end{equation*}
%
The type provider connects to the World Bank and obtains a list of countries at 
\emph{compile-time} and at \emph{edit-time} (when using auto-completion in an editor).
This means that the list is always up-to-date and we get a compile time error when accessing 
a country that no longer exists (a property discussed in Section~\ref{sec:tp-data}).

@@

Importing types for JavaScript libraries into the F\# type system has interesting implications, 
because the TypeScript language does not have the traditional type safety property \cite{ms-safets}.
We return to this topic in Section~\ref{sec:tp-lang}. Next, we generate checkboxes that appear on the 
right in Figure~\ref{fig:wb}:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{jQuery}~\ident{command}~=~\ident{j.jQuery.Invoke}(\ident{command}) 
 \\[0.5em]
 \kvd{let}~\ident{infos}~=~\ident{countries}~|\rang~\ident{List.map}~(\kvd{fun}~\ident{country}~\rightarrow \\
 \quad \kvd{let}~\ident{inp}~=~\ident{jQuery}(\str{<input>})\ident{.attr}(\str{type}, \str{checkbox}) \\
 \quad \ident{jQuery}(\str{\#panel})\ident{.append}(\ident{inp})\ident{.append}(\ident{country.Name}) \\
 \quad \ident{country.Name}, \ident{country.Indicators}, \ident{el}) \\
\end{array}
\end{equation*}
%
To manipulate the DOM (Document Object Model), we are using the jQuery library in a way that is
very similar to code that one would write in JavaScript. We define a helper function \ident{jQuery}
(hiding some of the complexities of the mapping) and use it to create the \str{<input>} element and
specify its attributes. Note that members like \ident{append} and \ident{attr} are standard 
jQuery patterns. The compiler sees them as ordinary object members. When writing code using F\# editors
based on the F\# Compiler Service \cite{fsharp-fcs}, they also appear in the auto-complete list.

@@

The main part of the sample program is a function \ident{render} that asynchronously fetches 
data for selected countries and generates a chart. To keep the code simple, we iterate over the 
\ident{infos} list from the previous section and load data for countries one by one:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{render}~()~=~\ident{async}~\{ \\
 \quad \kvd{let}~\ident{head}~=~\str{School enrollment, tertiary (\% gross)} \\
 \quad \kvd{let}~\ident{o}~=~\ident{h.HighchartsOptions}() \\
 \quad \ident{o.chart} \leftarrow \ident{h.HighchartsChartOptions}(\ident{renderTo}=\str{plc}) \\
 \quad \ident{o.title} \leftarrow \ident{h.HighchartsTitleOptions}(\ident{text}~=~\ident{head}) \\
 \quad \ident{o.series} \leftarrow \lbrack| ~ |\rbrack
 \\[0.5em]   
 \quad \kvd{for}~\ident{name},~\ident{ind},~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do}\\
 \qquad   \kvd{if}~\ident{unbox}\langl \ident{bool} \rangl~(\ident{check.is}(\str{:checked}))~\kvd{then} \\
 \qquad\quad     \kvd{let!}~\ident{v}~=~\ident{ind.}\lident{School enrollment, tertiary (\% gross)} \\
 \qquad\quad     \kvd{let}~\ident{data}~=~\ident{vals}~
                   |\rang~\ident{Seq.map}~(\kvd{fun}~(\ident{k}, \ident{v})~\rightarrow \lbrack|~\ident{number~k};~\ident{number~v} ~|\rbrack)~
                   |\rang~\ident{Array.ofSeq} \\
 \qquad\quad     \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name})) ~\}
\end{array}
\end{equation*}
%
Although the function looks like ordinary code, it is wrapped in the $\ident{async}~\{\ldots\}$ 
block, which is an F\# computation expression \cite{fsharp-zoo}. The F\# compiler performs 
de-sugaring similar to the CPS transformation and interprets keywords such as \kvd{let!} and 
\kvd{for} using special operations (monadic bind and other). The \ident{async} identifier 
determines that we are writing asynchronous workflow \cite{fsharp-async} that makes it 
possible to include non-blocking calls in the block.

@@

\subsection{Learning from the case study}

@@

In terms of programming language theory, type providers can be easily explained anecdotally. 
They change the starting point for a type-checking of a program as follows:
%
\begin{equation*}
\hspace{8em}
\begin{array}{rclll}
 \emptyset &\narrow{\vdash}& e : \tau 
    &\qquad& (\textit{classical, or closed-world})\\
 \pi~(\includegraphics[width=1em,trim=0 1.5em 0 0]{earth.png})  &\narrow{\vdash}& e : \tau
    &\qquad& (\textit{type providers, or open-world})
\end{array}
\end{equation*}
%
Rather than starting the type-checking with an empty context, we start type-checking with a
context containing a \emph{projection} of some information from the world. In this paper, we do not
attempt to formalize type providers, so we avoid the question of what, formally, is the world here.
However, this analogy is useful for understanding how type providers work and how they can go wrong:

@@

In summary, the World Bank type provider fits a specific niche that we were exploring in the case 
study. That is, when we want to create an information analysis that accesses specific data from
an information source. With rising popularity of data journalism and Open Government Data
initiatives, we argue that this is an important problem domain. Here, we first specify the country 
codes and then obtain the country objects.

@@

Type providers are often described as a technique that simplifies data access. The TypeScript
type provider shows that there is a broader range of uses, including integration between different
programming languages. Another example of type provider for language integration is the R
provider \cite{fsharp-rprovider} which imports packages and functions of the statistical 
environment R.

@@

\item Second, if the imported language has a more precise type system than F\#, the type 
  provider has to drop some of the information. This can be done safely for types in 
  contravariant positions. In the unsafe case, the runtime needs to perform dynamic checks 
  (Section~\ref{sec:conc-interop}).
\end{itemize}

@@

An interesting example of type provider for a language with a weaker type system is the
R provider \cite{fsharp-rprovider}. It solves the mismatch by using runtime reflection 
to discover R functions and their declared parameters. However, R provider also generates
additional function overloads that let the user specify additional parameters (not explicitly 
declared in the signature). To make such interoperability easier, we proposed additional 
escape mechanism \cite{fsharp-paramsdict} that provide usable syntax when static type information 
cannot be fully recovered.

@@

Using a type provider is similar to writing a code generator that turns TypeScript 
\textcolor{strclr}{\ttfamily d.ts} files into the above OCaml annotation. The main difference
is that type providers provide a smoother integration. Type providers do not produce any artefacts
(generated files) and can import entire repositories lazily.

@@

When we write code using type providers that provide erased types (both World Bank and TypeScript), 
then the erasure happens before a quotation is captured. The following listing shows an example. 
When we write code with explicit (or implicit) quotation containing a provided method (1), the actual 
quotation (2) contains just calls to the underlying runtime library:
%
\begin{equation*}
\begin{array}{llr}
 \langl @~\ident{data.}\lident{Czech Republic}~@\rangl &\hspace{10em}& (1)\\
 \langl @~\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE}) ~@\rangl  && (2)
\end{array}
\end{equation*}
%
Thus, the FunScript component only needs to provide a client-side implementation of the underlying
operations \ident{GetCountries}, \ident{GetCountry} and \ident{AsyncGetIndicator}. This is done in the
same way as mappings for standard .NET libraries. Under the cover, \ident{AsyncGetIndicator} invokes
an AJAX request to the World Bank API. We provide more details about asynchronous execution in 
Section~\ref{sec:js-async}.

@@

\begin{itemize}
\item Numeric types including \ident{int} and \ident{float} are mapped to JavaScript numbers.
  However, the floating-point arithmetic in JavaScript differs from the one in F\# and so this changes 
  the semantics. Similarly, compilation of integer operators requires additional work (for 
  example, $1/2$ is not an integer division in JavaScript).
